---
title: "STAT 5045 Exam 3"
author: "Tripp Bishop/STAT 5045"
date: "November 22, 2024"
date-format: "long"
format: pdf
margin: 2.54
fig-align: "center"
execute: 
  warning: false
  message: false
header-includes:
  \usepackage{amsmath}
---

```{r setup}
library(tidyverse)
library(kableExtra)

# clear out the environment
rm(list = ls())

# set all plots to use a customised version of the minimal theme.
base_plot <- theme_minimal() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(size=12),
    axis.title.x = element_text(size=10),
    axis.title.y = element_text(size=10),
    axis.text.x = element_text(size=8),
    axis.text.y = element_text(size=8)
  )

# now all plots will render with the same basic settings.
theme_set(base_plot)

```

## Question 1
A study in Florida was done to determine the primary food choice of alligators. 
The primary choices were `Fish`, `Invertebrate`, `Reptile`, `Bird`, `Other`. The 
lake location and the size (longer or shorter than 2.3 metres) of the alligator 
was also recorded. The data can be obtained from the `MLGdata` package in R. Use 
`Fish` as the reference for `foodchoice`. Use `George` as the reference for 
`lake`. Use $<= 2.3$ as the reference for `size`. $\alpha=0.05$.

```{r question 1}
gators <- MLGdata::Alligators

glimpse(gators)
levels(gators$foodchoice)
gators$foodchoice <- relevel(gators$foodchoice, ref = "Fish")
```


(a) Provide the model with an interaction. Explain all of the terms and 
subscripts.


$$
\begin{aligned}

\end{aligned}
$$



(b) Fit the model and test the interaction. Interpret the test result.

```{r question 1b}

```


(c) Based on the results of part (b) select a final model. Provide any required
additional test for the final model and interpret each test.



(d) Completely interpret a change in the size of an alligator using your final 
model.


## Question 2
Students from an Australian university conducted an experiment to determine
which type of sandwich attracts the most ants. The data can be obtained from the
`MGLdata` package in R.

The variables in the data are:

- **Bread** *1 = rye, 2 = wheatmeal, 3 = multigrain, 4 = white*
- **Filling** *1 = Vegemite, 2 = peanut butter, 3 = ham and pickles*
- **Butter** *1 = yes, 2 = no*
- **Ant_count** *number of captured ants*
- **Order** *order of the experiment*

(a) Read in the data. Change the categorical variables from numeric to factor
with the appropriate labels. Provide your R code.

```{r question 2a}
ants <- MLGdata::Ants

glimpse(ants)

ants <- ants |> 
  mutate(
    Bread = fct_recode(as_factor(Bread), 
                       "rye"="1", 
                       "wheatmeal"="2", 
                       "multigrain"="3",
                       "white"="4"),
    Filling = fct_recode(as_factor(Filling),
                         "Vegemite" = "1",
                         "peanut butter" = "2",
                         "ham and pickles" = "3"),
    Butter = fct_recode(as_factor(Butter),
                        "yes" = "1",
                        "no" = "-1")
  )

str(ants)
```

(b) Use the `xtabs` function to create a table of ant counts by `Bread` and 
`Filling`, ignore `Butter`. Provide the table. Test for an association between
`Bread` and `Filling`. Interpret the test result.

```{r question 2 part b}
xtabs(Ant_count ~ Bread + Filling, data = ants)
```

(c) Use the `xtabs` function to create a table of ant counts by `Bread` and 
`Filling` and by `Butter`. You should have a table of `Bread` by `Filling` for 
`no Butter` and another table of `Bread` by `Filling` for `yes Butter`.

```{r question 2 part c}
xtabs(Ant_count ~ Bread + Filling + Butter, data = ants)
```

(i) For the `no Butter` table, test the association between `Bread` and 
`Filling`. Interpret the test result.

(ii) For the `yes Butter` table, test the association between `Bread` and 
`Filling`. Interpret the test result.


## Question 3
A study was done to determine the lethality of gaseous carbon disulphide on
adult flour beetles. The beetles were exposed for 5 hours at different dosage
levels. The data can be obtained from the MLGdata package in R. $\alpha = 0.1$.

The variables in the data are:

- **num** *number of beetles exposed*
- **uccisi** *number of beetles that died*
_ **logdose** *concentration of carbon disulphide (mg per litre) in logarithmic scale*.


```{r question 3}
beetles <- MLGdata::Beetles
glimpse(beetles)
```




(a) Fit a model with `logdose` as the predictor. Provide the R code and the 
estimated regression model.

```{r question 3a}

```

(b) Provide R code and result for the test of including `logdose` in the model
or not. Interpret the test results.

```{r question 3 part b}

```

(c) For a 0.01 increase in the `logdose`, provide an estimate in the change of 
the odds with a CI. Interpret both the estimate and the CI.

```{r question 3 part c}

```


## Question 4
A study in Florida was done to determine the primary food choice of alligators.
The data in the file `allg8rLength.csv` contains the alligators only from Lake
George. The primary food choices are `Invertebrates`, `Fish`, and `Other`. The
length of the alligators is given in metres. Use `Fish` as the reference for
`foodchoice`.

```{r question 4}
gators <- read_csv("allg8rLength.csv")
glimpse(gators)
gators <- gators |> 
  mutate(
    foodChoice  = fct_relevel(as_factor(foodChoice), "Fish", after = 0),
    sex = fct_recode(as_factor(sex),
                     "Male" = "M",
                     "Female" = "F")
  )

levels(gators$foodChoice)

```




(a) Fit a model with `length`, `sex`, and the interaction. Provide the R code 
and the estimated regression model.

```{r question 4a}

```


(b) Test the interaction and provide the interpretation of the test.

```{r question 4b}

```

(c) Based on the results of the test of the interaction in part (b), select a 
final model. Provide the R code showing the final model and estimated 
coefficients. Using the estimated coefficients, fully interpret your final 
model. If length is in your final model, use a 1 decimetre (0.1 metre) increase
for the interpretation.

```{r question 4c}

```


## Question 5
Using the ant study of problem (2), compare the conclusions of parts (b) and 
(c). If the conclusions are the same, explain why. If the conclusions are
different, explain why. Do NOT repeat the conclusions, just explain why they are
similar or different.


## Question 6
Use the `wheat.csv` data to fit both nominal response and proportional odds
regression models to predict wheat kernel health. Both models should use the
variables `class`, `density`, `hardness`, `size`, `weight`, and `moisture`
*without interaction* to predict health status. The proportional odds model
should order then health status responses as `Scab`<`Sprout`<`Healthy`. Do not 
show the models you obtain, but show your code.

Now, consider the accuracy of these models. How well do each of these models
predict the different health statuses when compared to the actual status
reported in the data. For example, perhaps the proportional odds model correctly 
classified 80% of the Healthy wheat kernels. Report the overall accuracy for 
each health level. results should be reported in an easy to ready table with a 
clear description of the table contents.

One is clearly more accurate than the other, both overall and per health level.
Please give your understanding of why this is the case and provide at least one 
reason someone might want to use the less accurate model.

```{r question 6}

```


## Bonus Question
Using the alligator data from question 4 (`allg8rLength.csv`), produce a
probability plot based on the nominal response regression model in which the 
variable `length` only is used to predict food choice. Describe what this plot 
suggests about alligator length and food choice. Would a proportional odds model
be a good choice for this data? Why or why not?

```{r bonus question}

```


